---
title: "687 Descriptives 09.15.2025"
format: pdf
editor: visual
---

## 687 Descriptives

From the syllabus: "presenting simple (yet appropriate) descriptive statistics for each of the variables."

### Installing Necessary Packages and Reading in the Data

```{r}
library(dplyr)
library(tidyr)
library(ggridges)
library(gtsummary)
library(ggplot2)
#df_wide <- read.csv("cog_clean_wide.09.14.csv")
df_long <- read.csv("cog_clean_long.09.15.csv") # by far most updated data set
```

## Dealing with Missing Values

### Count Ns

```{r}
# just make sure certian variables are re-coded corretly
df_long <- df_long %>%
  mutate(SLFMEM = as.numeric(SLFMEM))

df_long <- df_long %>%
  mutate(SCHLYRS = as.numeric(SCHLYRS))

df_long <- df_long %>%
  mutate(DEGREE = as.factor(DEGREE))

df_long <- df_long %>%
  mutate(GENDER = as.factor(GENDER))

df_long <- df_long %>%
  mutate(RACE = as.factor(RACE))

df_long <- df_long %>%
  mutate(IWMODE = as.factor(IWMODE))

df_long <- df_long %>%
  mutate(IWLANG = as.factor(IWLANG))

# Find out how many respondents have and have not valid outcomes across ALL waves
result <- df_long %>%
  group_by(HHIDPN) %>%
  summarize(
    wave_count = n(),
    all_waves_present = n() == 11,
    all_valid_non_zero = all(!is.na(SLFMEM) & SLFMEM != 0),
    any_missing = any(is.na(SLFMEM)),
    any_zero = any(SLFMEM == 0, na.rm = TRUE)
  )

result %>%
  count(all_waves_present, all_valid_non_zero)

df_long %>%
  group_by(HHIDPN) %>%
  summarize(all_valid = all(!is.na(SLFMEM) & SLFMEM != 0)) %>%
  count(all_valid)
```

### Making Dataset for both only-Valid and only-Invalid

```{r}
# We start by finding HHIDPNs whose SLFMEM is 0 for at least 1 wave
zero_or_na_slfmem_ids <- df_long %>%
  group_by(HHIDPN) %>%
  summarize(has_zero_or_na = any(SLFMEM == 0, na.rm = TRUE) | any(is.na(SLFMEM))) %>%
  filter(has_zero_or_na == TRUE) %>%
  pull(HHIDPN)
```

```{r}
# Create a data set for those who have at least one 0 in any wave
df_all_zeros <- df_long %>% filter(HHIDPN %in% zero_or_na_slfmem_ids)

# Create a data set for those who have a valid value in any wave
df_long_filtered <- df_long %>% filter(!HHIDPN %in% zero_or_na_slfmem_ids)

# Thought I'd run this 
library(lme4)
library(nlme)
library(lmerTest)

mod1 = lmer(SLFMEM ~ 1 + (1|HHIDPN), REML = T, data = df_long_filtered) 
rand(mod1)
```

### Crosstab Summary for the Invalid data set (those who are dropped)

```{r}
# Step 1: Take one row per person (e.g., first wave)
person_level <- df_all_zeros %>%
  group_by(HHIDPN) %>%
  slice(1) %>%  # Takes the first row for each person
  ungroup()

# Step 2: Create characteristic variables (adjust based on your coding)
char <- person_level %>%
  mutate(
    Race = factor(RACE),  # Adjust factor levels as needed
    Gender = factor(GENDER),  # e.g., 1 = Male, 2 = Female
    Education = factor(DEGREE),  # Adjust levels
    SchoolYears = SCHLYRS,
    Interview_Language = factor(IWLANG),
    Interview_Mode = factor(IWMODE)
  ) %>%
  select(Race, Gender, Education, SchoolYears, Interview_Language, Interview_Mode)

# Step 3: Create summary table
char %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",  # For SCHLYRS if continuous
      all_categorical() ~ "{n} ({p}%)"
    ),
    label = list(
      Race = "Race",
      Gender = "Gender",
      Education = "Education Level",
      SchoolYears = "School Years",
      Interview_Language = "Interview Language",
      Interview_Mode = "Interview Mode"
    ),
    missing = "no"
  ) %>%
  modify_header(label = "**Characteristic**") %>%
  bold_labels()
```

## updated table 11.1

```{r}
person_level <- df_all_zeros %>%
  group_by(HHIDPN) %>%
  slice(1) %>%
  ungroup()

char <- person_level %>%
  mutate(
    Gender = case_when(
      is.na(GENDER) | GENDER == "" ~ "Unknown",
      tolower(GENDER) == "male" ~ "Male",
      tolower(GENDER) == "female" ~ "Female",
      TRUE ~ "Unknown"
    ) %>% factor(levels = c("Unknown", "Male", "Female")),
    Race = factor(RACE, levels = c(0, 1, 2, 7),
                  labels = c("Not Obtained", "White/Caucasian", 
                             "Black or African American", "Other")),
    SchoolYears = case_when(
      SCHLYRS == 0 ~ "No Formal Education",
      SCHLYRS >= 1 & SCHLYRS <= 11 ~ "1-11 Grades",
      SCHLYRS == 12 ~ "High School",
      SCHLYRS >= 13 & SCHLYRS <= 15 ~ "Some College",
      SCHLYRS == 16 ~ "College Graduate",
      SCHLYRS >= 17 ~ "Post College (17+ Years)",
      SCHLYRS == 99 | is.na(SCHLYRS) ~ "Missing",
      TRUE ~ "Missing"
    ),
    SchoolYears = factor(SchoolYears, levels = c(
      "No Formal Education", "1-11 Grades",
      "High School", "Some College",
      "College Graduate", "Post College (17+ Years)", "Missing"),
      ordered = TRUE),
    Interview_Language = factor(IWLANG, levels = c(0, 1, 2),
                               labels = c("Blank", "English", "Spanish")),
    Interview_Mode = case_when(
      IWMODE == 8 | is.na(IWMODE) ~ "Missing",
      IWMODE == 1 ~ "Face to Face",
      IWMODE == 2 ~ "Telephone",
      IWMODE == 3 ~ "Web",
      TRUE ~ "Missing"
    ) %>% factor(levels = c("Face to Face", "Telephone", "Web", "Missing"))
  ) %>%
  select(Race, Gender, SchoolYears, Interview_Language, Interview_Mode)

# View structure for troubleshooting
str(char)

# Final summary table
char %>%
  tbl_summary(
    statistic = list(all_categorical() ~ "{n} ({p}%)"),
    label = list(
      Race = "Race",
      Gender = "Gender",
      SchoolYears = "School Years",
      Interview_Language = "Interview Language",
      Interview_Mode = "Interview Mode"
    ),
    missing_text = "Missing",
    missing = "ifany"
  ) %>%
  modify_header(label = "**Characteristic**") %>%
  bold_labels()
```

```{r}
tab_race <- char %>% tabyl(Race)
tab_gender <- char %>% tabyl(Gender)
tab_schoolyears <- char %>% tabyl(SchoolYears)
tab_language <- char %>% tabyl(Interview_Language)
tab_mode <- char %>% tabyl(Interview_Mode)

# Each tabyl returns a data frame of counts and percents,
# You can print individually or bind into one table (with some cleaning for beautiful display):

# For illustration: show all together (not publication-ready, but clear)
list(
  Race = tab_race,
  Gender = tab_gender,
  SchoolYears = tab_schoolyears,
  Interview_Language = tab_language,
  Interview_Mode = tab_mode
)


```

## Variables

### Look at each demographic variable

```{r}
# 1) Number of all unique HHIDPNs in original dataset
original_unique <- length(unique(df_long$HHIDPN))
cat("1) Unique HHIDPNs in original dataset:", original_unique, "\n")

# 2) Number of all unique HHIDPNs in df_all_zeros
zeros_unique <- length(unique(df_all_zeros$HHIDPN))
cat("2) Unique HHIDPNs in all-zeros dataset:", zeros_unique, "\n")

# 3) Number of all unique HHIDPNs in filtered dataset  
filtered_unique <- length(unique(df_long_filtered$HHIDPN))
cat("3) Unique HHIDPNs in filtered dataset:", filtered_unique, "\n")

```

```{r}
# GENDER distribution
cat("GENDER distribution:\n")
cat("=== All Zeros Dataset ===\n")
table(df_all_zeros$GENDER) %>% print()
cat("=== Filtered Dataset ===\n")
table(df_long_filtered$GENDER) %>% print()


```

```{r}
# DEGREE distribution  
cat("\nDEGREE distribution:\n")
cat("=== All Zeros Dataset ===\n")
table(df_all_zeros$DEGREE) %>% print()
cat("=== Filtered Dataset ===\n")
table(df_long_filtered$DEGREE) %>% print()

# RACE distribution
cat("\nRACE distribution:\n")
cat("=== All Zeros Dataset ===\n")
table(df_all_zeros$RACE) %>% print()
cat("=== Filtered Dataset ===\n")
table(df_long_filtered$RACE) %>% print()

# IWLANG distribution
cat("\nIWLANG distribution:\n")
cat("=== All Zeros Dataset ===\n")
table(df_all_zeros$IWLANG) %>% print()
cat("=== Filtered Dataset ===\n")
table(df_long_filtered$IWLANG) %>% print()

# IWMODE distribution
cat("\nIWMODE distribution:\n")
cat("=== All Zeros Dataset ===\n")
table(df_all_zeros$IWMODE) %>% print()
cat("=== Filtered Dataset ===\n")
table(df_long_filtered$IWMODE) %>% print()
```

### Look at distribution of DV

#### Boxplot

```{r}
ggplot(df_long_filtered, aes(x = factor(wave), y = SLFMEM)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  labs(x = "Wave", y = "Dependent Variable", 
       title = "Distribution of DV by Wave") +
  theme_minimal()
```

#### density ridges

```{r}
ggplot(df_long_filtered, aes(x = SLFMEM, y = factor(wave), fill = factor(wave))) +
  geom_density_ridges(alpha = 0.7) +
  labs(x = "Dependent Variable", y = "Wave", 
       title = "Distribution of DV Across Waves") +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Combined box-plot

```{r}
ggplot(df_long_filtered, aes(x = factor(wave), y = SLFMEM)) +
  geom_jitter(alpha = 0.1, width = 0.2, color = "gray50") +  # Show individual points
  stat_summary(fun = mean, geom = "point", size = 3, color = "red") +  # Mean
  stat_summary(fun = mean, geom = "line", aes(group = 1), color = "red", size = 1) +  # Connect means
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, color = "red") +  # Error bars
  labs(x = "Wave", y = "Dependent Variable", 
       title = "Distribution of DV by Wave (with Mean ± SE)") +
  theme_minimal()

```

#### spaghetti plot with sample of 20

```{r}
# Sample a subset of individuals to avoid overplotting
set.seed(123)
sample_ids <- sample(unique(df_long_filtered$HHIDPN), 20)  # Adjust sample size

df_sample <- df_long_filtered %>%
  filter(HHIDPN %in% sample_ids)

mean_trajectory <- df_sample %>%
  group_by(wave) %>%
  summarise(mean_DV = mean(SLFMEM, na.rm = TRUE), .groups = "drop")

# Plot with unique colors for each person
ggplot(df_sample, aes(x = wave, y = SLFMEM, group = HHIDPN, color = factor(HHIDPN))) +
  geom_line(size = 1, alpha = 0.7) +  # Individual trajectories
  geom_point(size = 2, alpha = 0.7) +  # Add points on each line
  geom_line(data = mean_trajectory, aes(y = mean_DV, group = 1), 
            color = "black", size = 2, linetype = "dashed") +  # Mean trajectory
  geom_point(data = mean_trajectory, aes(y = mean_DV, group = 1), 
             color = "black", size = 4) +  # Mean points
  labs(x = "Wave", 
       y = "Dependent Variable", 
       title = "Individual Trajectories with Mean Trend (Black Dashed)",
       color = "Individual ID") +
  theme_minimal() +
  theme(legend.position = "right")

```

#### spaghetti plot by gender

```{r}
ggplot(df_long_filtered, aes(x = wave, y = SLFMEM, group = HHIDPN, color = GENDER)) +
  geom_line(alpha = 0.3) +
  labs(x = "Wave", y = "Dependent Variable", 
       title = "Individual Trajectories by Gender") +
  theme_minimal() +
  scale_color_manual(values = c("steelblue", "coral"))

```
